Citation Analysis of Computer Science

### 2019-11-18 Scientometrics/comp.tex
edits to latex and bibtex, mainly focusing on low-level writing and text formtting

In a historical study of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles), Salton and Bergmark made the key point that the scientific literature serves a rich source of information to study the structure and historical development of a field~\cite{salton_citation_1979}. They described the global structure of computer science as consisting of three main subareas (i) theoretical foundations, e.g., theory of computation (ii) hardware and computer systems, e.g., architecture (iii) software, e.g, programming systems.  Related areas acknowledged were  (a) mathematics of computing, e.g., numerical analysis (b) special software topics, e.g., operating systems (c) data management and database systems (d) methodologies valid for multiple applications, e.g., algebraic manipulation (e) computer applications, e.g., computer graphics (f) nontechnical aspects, e.g., computer education.

@@

In extending the work of Salton and Bergmark to the 21st century, we sought to combine journal and article based classifications in a scalable manner. Consequently, we used a whole text search method to match records from two major data sources. (i) the dblp bibliography- a reference bibliography for computer science~\cite{dblp_ref} (ii) the journal based Scopus bibliography, which more comprehensively covers the physical sciences, life sciences, social sciences, and health sciences. The underlying assumption is that a record in dblp qualifies as computer science. Therefore, we cross-matched dblp publications to Scopus to  enable identification of linked publications across multiple disciplines.

Using both direct citations and co-citations as the basis for clustering, we used articles from from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the the All Science Journal Classification (ASJC) developed and maintained by Scopus as well as to the CCS classification through a combination of automated and manual procedures. We describe below, high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences, and Life Sciences.

\section{Materials and Methods}

\emph{Data} A stable release of the dblp computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the xml source file and loaded into a PostgreSQL database. We have previously parsed, the Scopus dataset of over 72,000,000 publications into a custom schema as part of a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the dblp dataset, were matched to Scopus identifiers using PostgreSQL's full text search feature applied to document titles. The process involves parsing all Scopus document titles into tokens, reducing the tokens into lexemes and returning vectors containing these lexemes. The vectors are then matched with corresponding text search vectors extracted from DBLP document titles. A rank is returned based on the frequency of matching lexemes. Only recordxs in dblp with DOIs (4,291,130) were considered and only matches with a normalized rank of $>$ 0.99 were accepted, roughly equivalent to `at least 11 matching lexemes independent of order and vector length'. As an example, the following query returns a normalized rank of 0.994.
\begin{lstlisting}
SELECT ts_rank(to_tsvector(`english', `one two three 
four five six seven eight nine ten eleven'), 
plainto_tsquery(`english', `one four three two five 
six seven eight nine ten eleven')); 
\end{lstlisting}

This procedure resulted in a dataset of 634,100 dblp publications with Scopus identifiers. Cited references were extracted from Scopus, resulting in a total of 3,960,503 nodes (publications and references) and 11,992,387 edges that represent citations of these references and publications within the dataset. This dataset is referred to as \emph{comp}.\par

\emph{Clustering by Direct Citation} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}. We downloaded v1.2 from the authors' website at `cs.utexas.edu'. The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus,  We varied the number of clusters to be formed between 10 and 50 clusters. At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}. [@sitaram please add the number of nodes in the Graclus 20 clusters].

\emph{Clustering by Co-Citation} To build clusters using co-citations, we used variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges are dissolved) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size is put back into the process and a higher threshold applied to break such clusters.

@@

We applied variable clustering with initial parameters of threshold ($t$) = 0.5 (normalized co-citation frequency, see Methods), increment $i = 0.1$, and maximum cluster size, $mcs = 100$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9<= t<= 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 6,822 clusters containing 35,485 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 3,851. Pairwise clustering was then performed on these clusters to generate super-clusters. After 9 rounds of pairwise clustering, 13 clusters were generated resulting from agglomeration of 3,851 clusters

\label{sec:1}
Text with citations \cite{RefB} and \cite{RefJ}.

\section{Results}
\label{sec:2}

For the purpose of this study, we defined, the computer science literature as a working set of all articles in the dblp bibliography published in 1996-2015, that had a digital object identifier (DOI), and could be matched to article identifiers in the Scopus bibliography. The process of selection and matching resulted in a dataset of 634,100 publications (Materials and Methods). We constructed article level clusters of the computer science literature for the years 1996-2015, at a sufficiently high level of aggregation to be comparable to disciplinary classfication by journal. Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using a spectral clustering technique, Graclus, that has previously been applied to citation data~\cite{graclus_2007, subelj_clustering_2016}.  \emph{@sitaram please confirm the number of pubs that went into Graclus. I'm assuming it was 634,100?}. To place these clusters in context, we identified that accounted for at least 10\% of the publications in each cluster and observed 31 of the 334 minor subject areas in the Scopus ASJC classification. (which consists of 5 subject areas, 27 major subject areas, and 334 minor subject areas).

@@

Examining these data from the perspective of the generated clusters (Fig. 1 abscissa), interactions with the Life Sciences were evident in cluster 17 (Biochemistry, Computational Mathematics, Computational Theory and Mathematics, Computer Science Applications, and Molecular Biology) and cluster 10 (Artificial Intelligence, Biomedical Engineering, Cognitive Neuroscience, Computer Science Applications, and Neurology). While acknowledging previously reported inaccuracies in journal-based classification\emph{vide supra}, these results strongly suggest that the actual structure of communication within the broad area of computer science is at variance with journal-based classification, also noted by Waltman and van Eck, to wit, "...an indication that traditional disciplines such as those just mentioned only partly reflect the actual organization of today’s scientific research."~\cite{waltman_new_2012}. The aphorism of Porter, Cohen, Roessner, and Perreault (2007), `The discrepancies between disciplinary roots and research realities loom large'~\cite{porter_measuring_2007} also invokes ityself. Secondly, that a correlation exists between broad journal-based classification and article clusters in terms of subject matter. Thirdly, that some areas of computer science are relatively ubiquitous. Fourth that interactions with fields outside computer science tend to be restricted to a single cluster or small numbers of clusters.

Cite Alan Porter on expanding universe versus myopic focus. Cite Traag, Waltman, Lariviere papers. Peter Sjögårde and Per Ahlgren, Loet Leydesdorff, Caroline S. Wagner and Lutz Bornmann, Discontinuities in citation relations among journals: self-organized criticality as a model of scientific revolutions and change, Scientometrics, 10.1007/s11192-018-2734-6, 116, 1, (623-644), (2018). Antonio Perianes-Rodriguez and Javier Ruiz-Castillo, A comparison of the Web of Science and publication-level classification systems of science, Journal of Informetrics, 10.1016/j.joi.2016.10.007, 11, 1, (32-45), (2017). Qi Wang and Ludo Waltman, Large-scale analysis of the accuracy of the journal classification systems of Web of Science and Scopus, Journal of Informetrics, 10.1016/j.joi.2016.02.003, 10, 2, (347-364), (2016). Henry Small, Kevin W. Boyack and Richard Klavans, Identifying emerging topics in science and technology, 
Crossref
Overall strategy:

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{scopus_dblp_graclus2.pdf}
% figure caption is below the figure
\caption{insert caption}
\label{heatmap}       % Give a unique label
\end{figure}

\begin{itemize}
\item Spectral clustering
\begin{enumerate}
\item Subset Scopus by COMP
\item Match COMP to dblp at high stringency -> reduced number of scps
 \item Harvest cited references  from Scopus
 \item Cluster articles by direct citation using Graclus to a number that meets simple optimality criteria and does not impose cognitive challenges
 \item Reconcile cluster contents to journal classification for the purpose of referencing
 \item Infer inter-cluster relationships using direct citations, co-citations, and bibliographic coupling
\end{enumerate}
\item Co-citation based clustering
\begin{enumerate}
\item Stratify by year after subsetting by COMP and restricting to article and cp only
\item Apply fractional citation counting and identify highly-cited papers
\item Calculate normalized co-citations
\item Cluster by variable clustering method of Small
\item Evaluate major co-cited pairs
\end{enumerate}
\end{itemize}

@@

% Non-BibTeX users please use
\begin{thebibliography}{}
%
% and use \bibitem to create references. Consult the Instructions
% for authors for reference list style.
%
\bibitem{RefJ}
% Format for Journal Reference
Author, Article title, Journal, Volume, page numbers (year)
% Format for books
\bibitem{RefB}
Author, Book title, page numbers. Publisher, place (year)
% etc
\end{thebibliography}

>>>>

In a historical study of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles), Salton and Bergmark made the key point that the scientific literature serves a rich source of information to study the structure and historical development of a field~\cite{salton_citation_1979}. They described the global structure of computer science as consisting of three main subareas 
(i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas acknowledged were  (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

@@

In extending the work of Salton and Bergmark to the 21st century, we sought to combine journal and article based classifications in a scalable manner. Consequently, we used a whole text search method to match records from two major data sources. (i) the DBLP bibliography- a reference bibliography for computer science~\cite{dblp_ref} and  (ii) the journal-based Scopus bibliography, which more comprehensively covers the physical sciences, life sciences, social sciences, and health sciences. The underlying assumption is that a record in DBLP qualifies as computer science. Therefore, we cross-matched DBLP publications to Scopus to  enable identification of linked publications across multiple disciplines.

Using both direct citations and co-citations as the basis for clustering, we used articles from from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the the All Science Journal Classification (ASJC) developed and maintained by Scopus as well as to the CCS classification through a combination of automated and manual procedures. We describe below the high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences and Life Sciences.

\section{Materials and Methods}
\label{sec:methods}

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the xml source file and loaded into a PostgreSQL database. We have previously parsed, the Scopus dataset of over 72,000,000 publications into a custom schema as part of a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset, were matched to Scopus identifiers using PostgreSQL's full text search feature applied to document titles. The process involves parsing all Scopus document titles into tokens, reducing the tokens into lexemes and returning vectors containing these lexemes. The vectors are then matched with corresponding text search vectors extracted from DBLP document titles. A rank is returned based on the frequency of matching lexemes. Only records in DBLP with DOIs (4,291,130) were considered and only matches with a normalized rank of greater than 0.99 were accepted, roughly equivalent to `at least 11 matching lexemes independent of order and vector length'. As an example, the following query returns a normalized rank of 0.994.
\begin{lstlisting}
SELECT ts_rank(to_tsvector(`english', `one two three 
four five six seven eight nine ten eleven'), 
plainto_tsquery(`english', `one four three two five 
six seven eight nine ten eleven')); 
\end{lstlisting}

This procedure resulted in a dataset of 634,100 DBLP publications with Scopus identifiers. Cited references were extracted from Scopus, resulting in a total of 3,960,503 nodes (publications and references) and 11,992,387 edges that represent citations of these references and publications within the dataset. This dataset is referred to as \emph{comp}.\par

\subsection{Clustering methods}

\paragraph{Clustering by Direct Citation} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}. 
\textcolor{blue}{Let's provide the exact URL as a citation}
We downloaded v1.2 from the authors' website at `cs.utexas.edu'. 
The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus,  We varied the number of clusters to be formed between 10 and 50 clusters. At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}. [@sitaram please add the number of nodes in the Graclus 20 clusters].

\paragraph{Clustering by Co-Citation} To build clusters using co-citations, we used variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges are dissolved) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size is put back into the process and a higher threshold applied to break such clusters.

@@

We applied variable clustering with initial parameters of threshold ($t$) = 0.5 (normalized co-citation frequency, see Methods), increment $i = 0.1$, and maximum cluster size, $mcs = 100$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 6,822 clusters containing 35,485 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 3,851. Pairwise clustering was then performed on these clusters to generate super-clusters. After 9 rounds of pairwise clustering, 13 clusters were generated resulting from agglomeration of 3,851 clusters

\section{Results}
\label{sec:results}

For the purpose of this study, we defined, the computer science literature as a working set of all articles in the DBLP bibliography published in 1996-2015, that had a digital object identifier (DOI), and could be matched to article identifiers in the Scopus bibliography. The process of selection and matching resulted in a dataset of 634,100 publications (Materials and Methods). We constructed article level clusters of the computer science literature for the years 1996-2015, at a sufficiently high level of aggregation to be comparable to disciplinary classification by journal. Clustering of publications is commonly accomplished through direct citation, bibliographic coupling, and co-citation with an argument being made for direct citation as the best approach to concentrate citation links~\cite{klavans_which_2017}. Accordingly, we first used direct citation links as the basis for cluster formation. A set of 20 clusters was generated using a spectral clustering technique, Graclus, that has previously been applied to citation data~\cite{graclus_2007, subelj_clustering_2016}.  \emph{@sitaram please confirm the number of pubs that went into Graclus. I'm assuming it was 634,100?}. To place these clusters in context, we identified that accounted for at least 10\% of the publications in each cluster and observed 31 of the 334 minor subject areas in the Scopus ASJC classification. (which consists of 5 subject areas, 27 major subject areas, and 334 minor subject areas).

@@

Examining these data from the perspective of the generated clusters (Fig. 1 abscissa), interactions with the Life Sciences were evident in cluster 17 (Biochemistry, Computational Mathematics, Computational Theory and Mathematics, Computer Science Applications, and Molecular Biology) and cluster 10 (Artificial Intelligence, Biomedical Engineering, Cognitive Neuroscience, Computer Science Applications, and Neurology). While acknowledging previously reported inaccuracies in journal-based classification \emph{vide supra}, these results strongly suggest that the actual structure of communication within the broad area of computer science is at variance with journal-based classification, also noted by Waltman and van Eck, to wit, ``...an indication that traditional disciplines such as those just mentioned only partly reflect the actual organization of today's scientific research."~\cite{waltman_new_2012}. The aphorism of Porter, Cohen, Roessner, and Perreault (2007), `The discrepancies between disciplinary roots and research realities loom large'~\cite{porter_measuring_2007} also invokes itself. Secondly, that a correlation exists between broad journal-based classification and article clusters in terms of subject matter. Thirdly, that some areas of computer science are relatively ubiquitous. Fourth that interactions with fields outside computer science tend to be restricted to a single cluster or small numbers of clusters.

Cite Alan Porter on expanding universe versus myopic focus. Cite Traag, Waltman, Lariviere papers. Peter Sjögårde and Per Ahlgren, Loet Leydesdorff, Caroline S. Wagner and Lutz Bornmann, Discontinuities in citation relations among journals: self-organized criticality as a model of scientific revolutions and change, Scientometrics, 10.1007/s11192-018-2734-6, 116, 1, (623-644), (2018). Antonio Perianes-Rodriguez and Javier Ruiz-Castillo, A comparison of the Web of Science and publication-level classification systems of science, Journal of Informetrics, 10.1016/j.joi.2016.10.007, 11, 1, (32-45), (2017). Qi Wang and Ludo Waltman, Large-scale analysis of the accuracy of the journal classification systems of Web of Science and Scopus, Journal of Informetrics, 10.1016/j.joi.2016.02.003, 10, 2, (347-364), (2016). Henry Small, Kevin W. Boyack and Richard Klavans, Identifying emerging topics in science and technology, 
Crossref

\textcolor{blue}{George, the rest needs to go into methods}

\paragraph{Overall strategy}

Here we describe the overall strategy used for the two clustering methods.\\
\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{scopus_dblp_graclus2.pdf}
% figure caption is below the figure
\caption{insert caption}
\label{heatmap}       % Give a unique label
\end{figure}

\noindent
 Spectral clustering:
\begin{enumerate}
\item Subset Scopus by COMP
\item Match COMP to DBLP at high stringency -> reduced number of scps
 \item Harvest cited references  from Scopus
 \item Cluster articles by direct citation using Graclus to a number that meets simple optimality criteria and does not impose cognitive challenges
 \item Reconcile cluster contents to journal classification for the purpose of referencing
 \item Infer inter-cluster relationships using direct citations, co-citations, and bibliographic coupling
\end{enumerate}

\noindent
Co-citation based clustering:
\begin{enumerate}
\item Stratify by year after subsetting by COMP and restricting to article and cp only
\item Apply fractional citation counting and identify highly-cited papers
\item Calculate normalized co-citations
\item Cluster by variable clustering method of Small
\item Evaluate major co-cited pairs
\end{enumerate}

@@

% Non-BibTeX users please use

\clearpage
